For Finitary Induction-Induction, \\ Induction is Enough

### 2020-05-03 paper/universaliit.tex
stuff

The construction consists of the following steps:

\begin{enumerate}
\item Definition of untyped syntax (as a family of inductive datatypes) together
  with typing judgments (as inductive relations on the untyped syntax), and
  construction of a model of the theory of IIT signatures from well-formed
  terms, denoted $\mathsf{S}$.
\item Construction of a morphism $\rec:\mathsf{S}\to M$ for arbitrary $M$, by:
  \begin{enumerate}
  \item defining a relation $\blank \sim \blank$ between the well-typed syntax
    and a given model. The idea is that given a syntactic context $\GG$ and a
    semantic context $\model{\GG}$ of the model $M$, we have
    $\GG \sim \model{\GG}$ if and only if $\rec\,\,\GG
    = \model{\GG}$, and similarly for types, terms, and substitutions;
    % semantic context $\model{\GG}$ of
    % the model $M$ if and only if $\GG$ relates to $\rec\,{\GG}$;
    % enjoyed by the initial morphism from the
    % syntax to a given model;
  \item showing that this relation is functional.
    \end{enumerate}
  \item Proving the uniqueness of this morphism by showing that any morphism
    $f:\mathsf{S}\to M$ satisfies the relation. For example, for any syntactic context
    $\GG$ we have $\GG\sim f\,\GG$.
      % TODO donner composantes sur les contextes
\end{enumerate}
The next sections detail each of these steps.
\subsection{Syntactic Model}

The goal is to define the syntactic model where contexts are pairs of a
precontext together with a well-formedness proof, and similarly for types, terms
and substitutions.

Crucially, we do not have conversion relations for typed syntax, nor do we need
to use quotients when giving the syntactic model. This is possible because there
are no $\beta$-rules in the theory of signatures. Hence, we consider only normal
terms in the untyped syntax, and define weakening and substitution by recursion.
Avoiding quotients is important for two reasons. First, it greatly simplifies
formalisation. Second, we aim to reduce IITs to a minimal feature set, and we
get a stronger result if we do not use quotients.

The next sections present the definition of the untyped syntax and the
associated typing judgments.

\subsubsection{Untyped syntax}
The untyped syntax is defined as the following inductive datatype. Variables are
modeled as de Bruijn indices, i.e.\ as natural numbers pointing to a position in
the context. We use the additional default constructor $\erru:\Tmu$ in case of
error (ill-scoped substitution). The typing judgments will not mention $\erru$.
The main interest of $\erru$ is that it behaves like a closed term (which the
theory of signatures lacks), in the sense that it is
invariant under substitution. This makes expected equalities about 
substitution true even in the ill-typed case,
thus reducing the number of hypotheses for the corresponding lemmas
(see Section~\ref{ss:exch-laws}).
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \Conu && : \Set \\
  & \Tyu  && : \Set \\
  & \Subu  && : \Set \\
  % & \Subpre  && : \Set \\
  & \Tmu  && : \Set \\
  & \emptyu && : \Conu \\
    & \epsilonu && : \Subu \\
  & \blank\extu\blank && : \Conu\ra\Tyu\ra\Conu \\
  & \blank\consu\blank && : \Subu\ra\Tmu\ra\Subu \\
  & \varu  && : \N \ra \Tmu \\
  & \rlap{(2) Universe} \\[0.5em]
  & \Uu && : \Tyu \\
  & \Elu && : \Tmu\ra\Tyu \\
  & \rlap{(3) Inductive parameters} \\[0.5em]
  & \Piu && : \Tmu\ra\Tyu\ra\Tyu \\
  & \blank\appu\blank && : \Tmu\ra\Tmu\ra \Tmu \\
  & \rlap{(4-5) Metatheoretic parameters} \\[0.5em]
  & \Pimu && : (T:\Set)\ra(T\ra\Tyu)\ra\Tyu \\
  & \Piiu && : (T:\Set)\ra(T\ra\Tmu)\ra\Tmu \\
  & \blank\appmu \blank && : \Tmu\ra(\alpha:T) \ra\Tmu\\
  & \rlap{(6) Default value} \\[0.5em]
  & \erru && : \Tmu
\end{alignat*}

@@

\subsubsection{Untyped weakening}

% Substitutions of types and terms are defined recursively.

Note that $(\Piu \,A\,B)[\sigma]$ should be defined as $\Piu
\,(A[\sigma])\,(B[\wkO\,\sigma \,\consu\, \varu\,\mathsf{0}])$, and thus we need to define
$\wkO$, the weakening of substitutions.  The basic idea is to increment the de
Bruijn indices of all the variables.  Actually, this is not so simple because of
the $\Piu$ type: we want to define $\wkO\,(\Piu\,A\,B)$ as the $\Pi$ type
of the weakening of $A$ and $B$, but here, $B$ must be weakened with respect to
the second last variable of the context, rather than the last one.  For this
reason, we need to generalise the weakening as occuring anywhere in the context.
\begin{alignat*}{5}
  & \wk_n && :  \Tyu\ra\Tyu \\
  & \wk_n && :  \Tmu\ra\Tmu \\
  % & \wk && : \N \ra \N\ra\N \\
  & \wkO && : \Subu\ra \Subu
  \end{alignat*}
  The natural number $n$ specifies at which position of the context the
  weakening occurs.
  Here, $\wkO$ weakens with respect to the last variable.

Later, in Section~\ref{ss:weaken-typing}, we show that weakening preserves
typing. Stating a typing rule for this operation requires weakening at the
middle of a context. We consider pairs of untyped contexts, which should be
thought of as a splitting of a context at some position. The full context is
recovered by merging the two components:
% (which appears in the typing rule of a weakening occuring in the middle of a context),
% and the length $\length{\GG}$ of a
% context $\GG$:
\begin{alignat*}{5}
  & \blank \merge \blank && :  \Conu\ra\Conu\ra\Conu \\
  & \GG \merge \cdot && :=  \GG \\
  & \GG \merge (\GD \extu A) && :=  (\GG \merge \GD)\extu A
\end{alignat*}
We think of the second context as a telescope over the first one. We also define
weakening for telescopes, which will be used to give typing rules for telescopes
in Section \ref{sec:typing_judgments}:
\begin{alignat*}{5}
  & \wkO && : \Conu\ra \Conu
  \\
  & \wkO\,\emptyu && := \emptyu
  \\
  & \wkO\,(\GD\,\extu A) && := \wkO\,\GD\,\extu \wk_{\length{\GD}}\,A
  \end{alignat*}
$\length{\GG}$ denotes the length of a the context $\GG$.

\subsubsection{Untyped substitution}
  We define single substitution by recursion on presyntax:
  % & \rlap{(1) Substitution calculus} \\[0.5em]
\begin{alignat*}{5}
  & \blank[\blank := \blank] && : \Tyu\ra\N\ra\Tmu\ra\Tyu \\
  & \blank[\blank := \blank] && : \Tmu\ra\N\ra\Tmu\ra\Tmu
  % \\
  % & \blank[\blank := \blank] && : \N\ra\N\ra\Tmu\ra\Tmu
  \end{alignat*}
  This is enough to define the typing judgments: indeed, the typing rule for application
  involves only a unary substitution.

  However, to construct the initial model of the QIIT, we need to define
  the full substitution calculus:
\begin{alignat*}{5}
  & \blank[\blank] && : \Tyu\ra\Subu\ra\Tyu \\
  & \blank[\blank] && : \Tmu\ra\Subu\ra\Tmu \\
  & \blank \circ \blank && : \Subu\ra\Subu\ra \Subu
  \end{alignat*}
These can be defined either by iterating unary substitutions, or by recursion
on untyped syntax: the two ways yield provably equal definitions. In the
following, we assume that it is defined by recursion. We also make use of the
following definition:
  \begin{alignat*}{5}
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  & \keep && : \Subu \ra \Subu \\
   &&& := \lambda \sigma. (\varu\, \mathsf{0} \,\consu\, \wkO\,\sigma)
  \end{alignat*}
  The idea is that if $\sigma$ is a substitution between contexts $\GG$ and
  $\GD$, then $\keep\,\sigma$ is a substitution between contexts $\GG\ext A[\sigma]$
  and $\GD\ext A$ for any type $A$. This occurs when defining
  $(\Piu\,A\,B)[\sigma]$ as $\Piu\,(A[\sigma])\,(B[\keep \,\sigma])$.

  We can define the identity substitution on a context $\GG$ as
  follows, where $\keep^{\length{\GG}}$ is $\keep$ iterated
  $\length{\GG}$ times:
\begin{alignat*}{5}
  & \idu && : \Conu \ra \Subu \\
  & && := \lambda \GG.\keep^{\length{\GG}} \epsilonu
  \end{alignat*}

  \subsubsection{Exchange laws: weakening and substitution}
  \label{ss:exch-laws}
  Many lemmas for types and terms are shown by induction on the untyped syntax.
  Below, $Z$ denotes either a term or a type.
\begin{alignat*}{5}
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  & \wk\mhyphen\wk && : \wk_{n+p+1} (\wk_n\,Z) = \wk_n(\wk_{n+p}\,Z)\\
  % n-subZ-wkZ : ∀ n A z → (lifZZ n A) [ n ↦ z ]Z  ≡ A
  & \wk_n[n] && : (\wk_{n}\,Z)[ n := z] = Z \\
  % lifZ-l-subV : ∀ n p u x → lifZZ (n + p) (x [ n ↦ u ]V) ≡ (lifZV (S (n + p)) x) [ n ↦ (lifZZ p u) ]V
  & \wk_+[] && : (\wk_{n+p+1}\,Z)[ n := \wk_p\,u] = \wk_{n+p}\,(Z[n := u]) \\
  % l-subV-lifZV : ∀ Δ u n x → (lifZV n x) [ (S (n + Δ)) ↦ u ]V  ≡ lifZZ n (x [ (n + Δ) ↦ u ]V)
  & \wk[+] && : (\wk_{n}\,Z)[ n+p+1 :=  u] = \wk_{n}\,(Z[n+p := u]) \\
  % l-subV-l-subV : ∀ n p z u x →  (x [ n ↦ u ]V) [ (n + p) ↦ z ]Z  ≡ (x [ (S (n + p)) ↦ z ]V) [ n ↦ (u [ p ↦ z ]Z) ]Z
  & [][+] && : Z[n :=u][n+p:=z] = Z[n+p+1:=z][n:= (u[p:=z])]
  \end{alignat*}
  Below, $\keep^n$ denotes $\keep$ iterated $n$ times.
\begin{alignat*}{5}
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  % liftV=wkS
  & [\keep^n\mhyphen \wkO] && : Z[\keep^n\,(\wkO\,\sigma)] = \wk_n (Z[\keep^n\,\sigma])
  \\
  % lifZₛV : ∀ n xp σp Zp → (lifZV n xp [ iZer n keep (Zp ∷ σp) ]V) ≡ (xp [ iZer n keep σp ]V)
  & \wk_n[\keep^n\mhyphen\cons]  & &
  :
  (\wk_n\,Z) [\keep^n\, (u \consu \sigma)] = Z [ \keep^n \sigma]
  \\
  % l-sub[]V : ∀ n x z σ
    % ((x [ n := z ]V) [ iZer n keep σ ]Z) = (x [ iZer (S n) keep σ ]V) [ n := (z [ σ ]Z) ]Z
   & [:=][\keep] && : Z [ n:= u][\keep^n\,\sigma] =  Z[\keep^{n+1}\,\sigma][n := u[\sigma]]
  \end{alignat*}
As particular cases for $n=0$, we get
\begin{alignat*}{5}
  & {}{\circ}{}\wkO & & : \sigma \circ (\wkO \tau) = \wkO (\sigma \circ \tau) \\
  & \wkO{}{\circ}{}, && : \wkO\,\sigma \circ (t \consu \tau ) = \sigma \circ \tau \\
  & [\wkO] && : t[\wkO\,\sigma] = \wk_0 (t[\sigma]) \\
  & \wk_0[\cons] && : (\wk_0 Z)[u \consu \sigma] = Z[\sigma] \\
  & [0:=][] && : Z [ 0:= u][ \sigma] =  Z[\keep\,\sigma][0 := u[\sigma]]
\end{alignat*}
Finally, we show the following:
\begin{alignat*}{5}
  & [][] && : Z[\sigma][\tau] = Z[\sigma\circ\tau]
  \\
  & \ass && : (\sigma\circ \delta)\circ\tau = \sigma\circ (\delta\circ\tau)
  \end{alignat*}
We defer laws for identity substitutions after the definition of the typing
judgments, as the proofs require that some inputs are well-typed.

@@

  \subsubsection{Typing judgments}
  \label{sec:typing_judgments}
  The typing judgments are defined as the following inductive datatype indexed over the
  untyped syntax:
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \blank \vdash && : \Conu\ra \Set \\
  & \blank \vdash \blank  && : \Conu\ra \Tyu\ra\Set \\
  & \blank \vdash \blank \invar \blank  && : \Conu\ra \N \ra \Tyu\ra \Set \\
  & \blank \vdash \blank \in \blank  && : \Conu\ra \Tmu\ra \Tyu\ra \Set \\
  & \blank \vdash \blank \Rightarrow \blank  && : \Conu\ra \Subu \ra \Conu\ra \Set \\
  % & \Subpre  && : \Set \\
  & \emptyw && : \emptyu \vdash \\
  & \epsilonw && : \GG\vdash \epsilonu \Rightarrow \emptyu \\
  & \blank\extw\blank && : ( \GG\vdash )  \ra  (\GG\vdash A)  \ra
  \GG\extu A \vdash \\
  & \consw && :
    (\GD \vdash) \ra
    (\GG\vdash \sigma \Rightarrow\GD)\ra
    ( \GD \vdash A) \ra
    ( \GG \vdash t \in A[\sigma]) \ra
    \GG \vdash t \consu \sigma \Rightarrow \GD\extu A
   \\
  & \varw  && : (\GG \vdash n \invar A) \ra \GG \vdash \varu n \in A \\
  & \vzw  && : (\GG\vdash)\ra(\GG\vdash A)\ra\GG \extu  A \vdash \mathsf{0} \invar \wku \, A \\
  & \vsw  && : (\GG\vdash)\ra(\GG\vdash A)\ra(\GG \vdash n \invar A) \ra (\GG \vdash B) \ra \GG \extu  B
  \vdash \mathsf{S}\,n \invar \wku \, A \\
  & \rlap{(2) Universe} \\[0.5em]
  & \Uw && : (\GG \vdash)\ra \GG \vdash \Uu \\
  & \Elw && : (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra\GG \vdash \Elu\,a \\
  & \rlap{(3) Inductive parameters} \\[0.5em]
\  & \Piw && :
    (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra(\GG \extu
  \Elu\,a \vdash B)\ra\GG \vdash \Piu \, a \, B \\
  & \appw && :
    (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra(\GG \extu
    \Elu\,a \vdash B)
    \\
    & &&
    \ra(\GG \vdash t \in \Piu \, a \, B )
    \ra
    (\GG \vdash u \in \Elu\,a)
    \ra
    \GG \vdash t \appu u \in  B [0 := u] \\
  & \rlap{(4) Inductive parameters} \\[0.5em]
  & \Pimw && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\GG \vdash)\ra
    ((t : T)\ra\GG\vdash A\,t) \ra
    \GG \vdash \Pimu \,T\,A
    \\
    & \appmw  && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\GG \vdash)\ra
    ((t : T)\ra\GG\vdash A\,t)
    \\
    & &&
    \ra(\GG \vdash t \in \Pimu \,T\,A)
    \ra (u : T) \ra \GG \vdash t\,\appmu\,u \in A\,u
\end{alignat*}
There is possibility of redundancy in the arguments of the constructors. Here,
we are ``{paranoid}'', so that we get more inductive hypotheses when performing
recursion.

>>>>

We construct the theory of IIT signatures in the following steps:
\begin{enumerate}
\item We view the theory of signatures as a type theory, and we define
  its untyped syntax as mutual inductive types together with typing
  judgments given by inductive predicates on the untyped syntax. Then
  the syntax $\I : \SignAlg$ is constructed using those untyped terms
  for which the typing predicate holds.
\item We construct $\ll\blank\rr:\SignMor\,\I\,M$ for arbitrary
  $M : \SignAlg$, by:
  \begin{enumerate}
  \item defining a relation $\blank \sim \blank$ between the well-typed syntax
    and a given model. The idea is that given a syntactic context $\GG$ and a
    semantic context $\model{\GG}$ of the model $M$, we have
    $\GG \sim \model{\GG}$ if and only if $\ll\GG\rr
    = \model{\GG}$, and similarly for types, terms, and substitutions;
    % semantic context $\model{\GG}$ of
    % the model $M$ if and only if $\GG$ relates to $\rec\,{\GG}$;
    % enjoyed by the initial morphism from the
    % syntax to a given model;
  \item showing that this relation is functional.
    \end{enumerate}
  \item Proving the uniqueness of this morphism by showing that any morphism
    $f:\SignMor\,\I\,M$ satisfies the relation. For example, for any syntactic context
    $\GG$ we have $\GG\sim f\,\GG$.
      % TODO donner composantes sur les contextes
\end{enumerate}
The next sections detail each of these steps.

\subsection{Syntax}

The goal is to define the syntactic signature algebra where contexts
are pairs of a precontext together with a well-formedness proof, and
similarly for types, terms and substitutions.

Crucially, we do not have conversion relations for typed syntax, nor
do we need to use quotients when giving the syntactic model. This is
possible because there are no $\beta$-rules in the theory of
signatures. Hence, we consider only normal terms in the untyped
syntax, and define weakening and substitution by recursion.  Avoiding
quotients is important for two reasons. First, it greatly simplifies
formalisation. Second, we aim to reduce the theory of signatures only
to inductive types, thus making \Cref{TODO} stronger.

Now we present the definition of the untyped syntax and the
associated typing judgments.

\subsubsection{Untyped syntax and its properties}

\begin{definition}[Untyped syntax]
The untyped syntax is defined as the following inductive datatype.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\
  & \Conu && : \Set \\
  & \Tyu  && : \Set \\
  & \Subu  && : \Set \\
  % & \Subpre  && : \Set \\
  & \Tmu  && : \Set \\
  & \emptyu && : \Conu \\
    & \epsilonu && : \Subu \\
  & \blank\extu\blank && : \Conu\ra\Tyu\ra\Conu \\
  & \blank\consu\blank && : \Subu\ra\Tmu\ra\Subu \\
  & \varu  && : \N \ra \Tmu \\
  & \rlap{(2) Universe} \\
  & \Uu && : \Tyu \\
  & \Elu && : \Tmu\ra\Tyu \\
  & \rlap{(3) Inductive parameters} \\
  & \Piu && : \Tmu\ra\Tyu\ra\Tyu \\
  & \blank\appu\blank && : \Tmu\ra\Tmu\ra \Tmu \\
  & \rlap{(4) Metatheoretic parameters} \\
  & \Pimu && : (T:\Set)\ra(T\ra\Tyu)\ra\Tyu \\
  & \Piiu && : (T:\Set)\ra(T\ra\Tmu)\ra\Tmu \\
  & \blank\appmu \blank && : \Tmu\ra(\alpha:T) \ra\Tmu\\
  & \rlap{(5) Default value} \\
  & \erru && : \Tmu
\end{alignat*}
\end{definition}
Variables are modeled as de Bruijn indices, i.e.\ as natural numbers
pointing to a position in the context. We use the additional default
constructor $\erru:\Tmu$ in case of error (ill-scoped
substitution). The typing judgments will not mention $\erru$.  The
main interest of $\erru$ is that it behaves like a closed term (which
the theory of signatures lacks), in the sense that it is invariant
under substitution. This makes expected equalities about substitution
true even in the ill-typed case, thus reducing the number of
hypotheses for the corresponding lemmas (see
\Cref{lem:exch-laws}).

@@

We will define substitutions $\blank[\blank]$ of types and terms
recursively.

Note that $(\Piu \,A\,B)[\sigma]$ should be defined as
$\Piu \,(A[\sigma])\,(B[\wkO\,\sigma \,\consu\, \varu\,\mathsf{0}])$,
and thus we need to define $\wkO$, the weakening of substitutions.
The basic idea is to increment the de Bruijn indices of all the
variables.  Actually, this is not so simple because of the $\Piu$
type: we want to define $\wkO\,(\Piu\,A\,B)$ as the $\Pi$ type of the
weakening of $A$ and $B$, but here, $B$ must be weakened with respect
to the second last variable of the context, rather than the last one.
For this reason, we need to generalise the weakening as occuring
anywhere in the context.

\begin{definition}[Untyped weakening]
  We define untyped weaking recursively on terms by the following
  functions.
  \begin{alignat*}{5}
    & \wk_n && :  \Tyu\ra\Tyu \\
    & \wk_n && :  \Tmu\ra\Tmu \\
    % & \wk && : \N \ra \N\ra\N \\
    & \wkO && : \Subu\ra \Subu
  \end{alignat*}
  The natural number $n$ specifies at which position of the context the
  weakening occurs.
  Here, $\wkO$ weakens with respect to the last variable.
\end{definition}

Later, in \Cref{lem:weaken-typing}, we show that weakening preserves
typing.  Stating a typing rule for this operation requires weakening
at the middle of a context. This is why we define pairs of untyped
contexts, which should be thought of as a splitting of a context at
some position. We call the second context a telescope over the first
one.
% (which appears in the typing rule of a weakening occuring in the middle of a context),
% and the length $\length{\GG}$ of a
% context $\GG$:
\begin{definition}[Untyped telescopes]
  An untyped telescope is given simply by a $\Conu$.
\end{definition}
\begin{definition}[Merging of a context and a telescope]
\begin{alignat*}{5}
  & \blank \merge \blank && :  \Conu\ra\Conu\ra\Conu \\
  & \GG \merge \cdot && :=  \GG \\
  & \GG \merge (\GD \extu A) && :=  (\GG \merge \GD)\extu A
\end{alignat*}
\end{definition}
\begin{definition}[Weakening for telescopes]
Weakening for telescopes is defined pointwise. $\length{\GG}$ denotes
the length of the context $\GG$.
\begin{alignat*}{5}
  & \wkO && : \Conu\ra \Conu
  \\
  & \wkO\,\emptyu && := \emptyu
  \\
  & \wkO\,(\GD\,\extu A) && := \wkO\,\GD\,\extu \wk_{\length{\GD}}\,A
  \end{alignat*}
\end{definition}
This will be used to give typing rules for telescopes
in \Cref{def:typing_judgments}.

\begin{definition}[Untyped unary substitution]
  We define single substitution by recursion on the presyntax:
  % & \rlap{(1) Substitution calculus} \\[0.5em]
\begin{alignat*}{5}
  & \blank[\blank := \blank] && : \Tyu\ra\N\ra\Tmu\ra\Tyu \\
  & \blank[\blank := \blank] && : \Tmu\ra\N\ra\Tmu\ra\Tmu
  % \\
  % & \blank[\blank := \blank] && : \N\ra\N\ra\Tmu\ra\Tmu
\end{alignat*}
\end{definition}
  This is enough to state the typing judgments: indeed, the typing rule for application
  involves only a unary substitution.

  However, to construct the syntax as a signature algebra, we need to define
  parallel substitutions:
\begin{definition}[Untyped substitution calculus]
  \begin{alignat*}{5}
  & \blank[\blank] && : \Tyu\ra\Subu\ra\Tyu \\
  & \blank[\blank] && : \Tmu\ra\Subu\ra\Tmu \\
  & \blank \circ \blank && : \Subu\ra\Subu\ra \Subu
  \end{alignat*}
These can be defined either by iterating unary substitutions, or by recursion
on untyped syntax: the two ways yield provably equal definitions. In the
following, we assume that they are defined by recursion. We also make use of the
following definition:
  \begin{alignat*}{5}
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  & \keep && : \Subu \ra \Subu \\
   &&& := \lambda \sigma. (\wkO\,\sigma \,\consu\, \varu\, \mathsf{0})
  \end{alignat*}
  The idea is that if $\sigma$ is a substitution from $\GG$ to
  $\GD$, then $\keep\,\sigma$ is a substitution between contexts $\GG\ext A[\sigma]$
  and $\GD\ext A$ for any type $A$ where the last term is just a de Bruijn index $0$. This occurs when defining
  $(\Piu\,A\,B)[\sigma]$ as $\Piu\,(A[\sigma])\,(B[\keep \,\sigma])$.

  We define the identity substitution on a context $\GG$ as
  follows, where $\keep^{\length{\GG}}$ is $\keep$ iterated
  $\length{\GG}$ times:
\begin{alignat*}{5}
  & \idu && : \Conu \ra \Subu \\
  & && := \lambda \GG.\keep^{\length{\GG}} \epsilonu
  \end{alignat*}
\end{definition}

\begin{lemma}[Exchange laws for weakening and substitution]\label{lem:exch-laws}
  Below, $Z$ denotes either a term or a type and $\keep^n$ denotes
  the $n$ times iteration of $\keep$.
\begin{alignat*}{5}
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  & \wk\mhyphen\wk && : \wk_{n+p+1} (\wk_n\,Z) = \wk_n(\wk_{n+p}\,Z)\\
  % n-subZ-wkZ : ∀ n A z → (lifZZ n A) [ n ↦ z ]Z  ≡ A
  & \wk_n[n] && : (\wk_{n}\,Z)[ n := z] = Z \\
  % lifZ-l-subV : ∀ n p u x → lifZZ (n + p) (x [ n ↦ u ]V) ≡ (lifZV (S (n + p)) x) [ n ↦ (lifZZ p u) ]V
  & \wk_+[] && : (\wk_{n+p+1}\,Z)[ n := \wk_p\,u] = \wk_{n+p}\,(Z[n := u]) \\
  % l-subV-lifZV : ∀ Δ u n x → (lifZV n x) [ (S (n + Δ)) ↦ u ]V  ≡ lifZZ n (x [ (n + Δ) ↦ u ]V)
  & \wk[+] && : (\wk_{n}\,Z)[ n+p+1 :=  u] = \wk_{n}\,(Z[n+p := u]) \\
  % l-subV-l-subV : ∀ n p z u x →  (x [ n ↦ u ]V) [ (n + p) ↦ z ]Z  ≡ (x [ (S (n + p)) ↦ z ]V) [ n ↦ (u [ p ↦ z ]Z) ]Z
  & [][+] && : Z[n :=u][n+p:=z] = Z[n+p+1:=z][n:= (u[p:=z])] \\
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  % liftV=wkS
  & [\keep^n\mhyphen \wkO] && : Z[\keep^n\,(\wkO\,\sigma)] = \wk_n (Z[\keep^n\,\sigma])
  \\
  % lifZₛV : ∀ n xp σp Zp → (lifZV n xp [ iZer n keep (Zp ∷ σp) ]V) ≡ (xp [ iZer n keep σp ]V)
  & \wk_n[\keep^n\mhyphen\cons]  & &
  :
  (\wk_n\,Z) [\keep^n\, (\sigma \consu u)] = Z [ \keep^n \sigma]
  \\
  % l-sub[]V : ∀ n x z σ
    % ((x [ n := z ]V) [ iZer n keep σ ]Z) = (x [ iZer (S n) keep σ ]V) [ n := (z [ σ ]Z) ]Z
   & [:=][\keep] && : Z [ n:= u][\keep^n\,\sigma] =  Z[\keep^{n+1}\,\sigma][n := u[\sigma]]
  \end{alignat*}
\end{lemma}
\begin{proof}
By induction on the untyped syntax.
\end{proof}

\begin{corollary}
As particular cases for $n=0$, we get
\begin{alignat*}{5}
  & {}{\circ}{}\wkO & & : \sigma \circ (\wkO \tau) = \wkO (\sigma \circ \tau) \\
  & \wkO{}{\circ}{}, && : \wkO\,\sigma \circ (\tau \consu t) = \sigma \circ \tau \\
  & [\wkO] && : t[\wkO\,\sigma] = \wk_0 (t[\sigma]) \\
  & \wk_0[\cons] && : (\wk_0\,Z)[\sigma \consu u] = Z[\sigma] \\
  & [0:=][] && : Z [ 0:= u][\sigma] =  Z[\keep\,\sigma][0 := u[\sigma]]
\end{alignat*}
\end{corollary}

\begin{lemma}[Composition functor law and associativity]
\begin{alignat*}{5}
  & [][] && : Z[\sigma][\tau] = Z[\sigma\circ\tau]
  \\
  & \ass && : (\sigma\circ \delta)\circ\tau = \sigma\circ (\delta\circ\tau)
\end{alignat*}
\end{lemma}

We defer laws for identity substitutions after the definition of the typing
judgments, as the proofs require that some inputs are well-typed.

@@

\subsubsection{Typing judgments and their properties}

\begin{definition}[Typing judgements]\label{def:typing_judgments}
  The typing judgments are defined as the following inductive type indexed over the
  untyped syntax:
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \blank \vdash && : \Conu\ra \Set \\
  & \blank \vdash \blank  && : \Conu\ra \Tyu\ra\Set \\
  & \blank \vdash \blank \invar \blank  && : \Conu\ra \N \ra \Tyu\ra \Set \\
  & \blank \vdash \blank \in \blank  && : \Conu\ra \Tmu\ra \Tyu\ra \Set \\
  & \blank \vdash \blank \Rightarrow \blank  && : \Conu\ra \Subu \ra \Conu\ra \Set \\
  % & \Subpre  && : \Set \\
  & \emptyw && : \emptyu \vdash \\
  & \epsilonw && : \GG\vdash \epsilonu \Rightarrow \emptyu \\
  & \blank\extw\blank && : ( \GG\vdash )  \ra  (\GG\vdash A)  \ra
  \GG\extu A \vdash \\
  & \consw && :
    (\GD \vdash) \ra
    (\GG\vdash \sigma \Rightarrow\GD)\ra
    ( \GD \vdash A) \ra
    ( \GG \vdash t \in A[\sigma]) \ra
    \GG \vdash t \consu \sigma \Rightarrow \GD\extu A
   \\
  & \varw  && : (\GG \vdash n \invar A) \ra \GG \vdash \varu n \in A \\
  & \vzw  && : (\GG\vdash)\ra(\GG\vdash A)\ra\GG \extu  A \vdash \mathsf{0} \invar \wku \, A \\
  & \vsw  && : (\GG\vdash)\ra(\GG\vdash A)\ra(\GG \vdash n \invar A) \ra (\GG \vdash B) \ra \GG \extu  B
  \vdash \mathsf{S}\,n \invar \wku \, A \\
  & \rlap{(2) Universe} \\[0.5em]
  & \Uw && : (\GG \vdash)\ra \GG \vdash \Uu \\
  & \Elw && : (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra\GG \vdash \Elu\,a \\
  & \rlap{(3) Inductive parameters} \\[0.5em]
\  & \Piw && :
    (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra(\GG \extu
  \Elu\,a \vdash B)\ra\GG \vdash \Piu \, a \, B \\
  & \appw && :
    (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra(\GG \extu
    \Elu\,a \vdash B)
    \\
    & &&
    \ra(\GG \vdash t \in \Piu \, a \, B )
    \ra
    (\GG \vdash u \in \Elu\,a)
    \ra
    \GG \vdash t \appu u \in  B [0 := u] \\
  & \rlap{(4) Inductive parameters} \\[0.5em]
  & \Pimw && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\GG \vdash)\ra
    ((t : T)\ra\GG\vdash A\,t) \ra
    \GG \vdash \Pimu \,T\,A
    \\
    & \appmw  && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\GG \vdash)\ra
    ((t : T)\ra\GG\vdash A\,t)
    \\
    & &&
    \ra(\GG \vdash t \in \Pimu \,T\,A)
    \ra (u : T) \ra \GG \vdash t\,\appmu\,u \in A\,u
  \end{alignat*}
\end{definition}
There is possibility of redundancy in the arguments of the constructors. Here,
we are ``{paranoid}'' (following nomenclature from xxx), so that we get more inductive hypotheses when performing
recursion.
